The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-29 paper.tex
Revisao Terra

\usepackage{verbatim}
\usepackage{xspace}
\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\code}[1] {{\small{\texttt{#1}}}}
\newcommand{\MM}[1] {\textcircled{\tiny{\textsf{#1}}}}

@@

An established alternative to \emph{C} in the field of embedded systems is the 
family of reactive synchronous languages \cite{rp.twelve}.
%
Two major styles of synchronous languages have evolved:
in the \emph{control}--\emph{imperative} style, programs are structured with 
control flow primitives, such as parallelism, repetition, and preemption;
in the \emph{dataflow}--\emph{declarative} style, programs can be seen as 
graphs of values, in which a change to a value is propagated through its 
dependencies without explicit programming.
%
Among the control-based languages, Esterel~\cite{esterel.ieee91} influenced a 
number of other embedded languages, such as \emph{Reactive-C}~\cite{rp.rc}, 
\emph{OSM}~\cite{wsn.osm}, \emph{Sync-C}~\cite{rp.synchc}, and 
\emph{PRET-C}~\cite{rp.pretc}.

@@

In this work, we present a formal semantics of a small synchronous kernel that 
represents a subset of \CEU covering these new functionalities.
%
We also present an implementation of \CEU with two backends:
one aiming for efficiency and interoperability with \emph{C},
and another based on a virtual machine that allows remote reprogramming.

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of constrained wireless sensor networks, developing a number of applications, 
protocols, and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\emph{C}, and observed a reduction in source code size (around 25\%) with a 
small increase in memory usage (around 5--10\% for ROM and 
RAM)~\cite{ceu.sensys13}.
%
Considering the \emph{VM} backend, a simple application that blinks three LEDs 
periodically occupies less than 100 bytes and can be completely transmitted in 
4 radio messages.
In contrast, without the preinstalled \emph{VM} with basic \emph{I/O} 
functionality, a \emph{C}-compiled version occupies more than 2 
kbytes~\cite{ceu.terra}.

The rest of the paper is organized as follows:
Section~\ref{sec.ceu} gives an overview of \CEU, exposing its fundamental 
differences to Esterel.
Section~\ref{sec.sem} presents a formal semantics for the control primitives of 
\CEU.
Section~\ref{sec.impl} presents the \emph{C} and \emph{VM} implementation 
backends.
Section~\ref{sec.related} discusses other synchronous languages targeting 
embedded systems.
Section~\ref{sec.conclusion} concludes the paper.

@@

To illustrate threats related to abortion, consider the unsafe example, which 
does not compile, in the left of Figure~\ref{lst.fin}, describing the state 
machine of a data collection protocol for sensor 
networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (line 1) 
represent the external interface of the protocol.
%
The protocol has to transmit a packet every minute, unless it receives a 
\code{RETRANSMIT} request to immediately re-transmit it, or a \code{STOP} 
request to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (line 3);
the other periodically transmits the packet (lines 5--18).
%
The periodic transmission is a loop that starts two other trails (lines 6--17):
one handles the immediate retransmission request (line 7);
the other transmits the packet%
\footnote{
The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
with external \emph{C} libraries explicit and are required in \CEU.
}
and waits for a confirmation (lines 9--16).
%
%
The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping, in accordance with the specification.
Note that the call to \code{\_send\_enqueue} is \emph{asynchronous}, handing to 
the radio driver a pointer to the lexically-scoped packet.
The driver makes the transmission in the background, holding the packet until 
it signals the application with the \code{SENDACK} to acknowledge completion.
%
At any time, the client may request a retransmission (line 7), which terminates 
the \code{par/or} (line 6), aborts the ongoing transmission (line 14, if not 
idle), and restarts the loop (line 5).
%
The client may also request to stop the whole protocol at any time (line 3).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver, which will possibly transmit 
corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

The unsafe example in the left of Figure~\ref{lst.fin} does not compile because 
\CEU tracks the interaction of \code{par/or} compositions with local variables 
and stateful \emph{C} functions (e.g., device drivers) in order to preserve 
safe abortion of trails~\cite{ceu.sensys13,ceu.mod15}.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
In fact, \CEU enforces the programmer to write a \emph{finalization} clause to 
accompany the stateful \emph{C} call.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}) executes automatically on external 
abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \emph{C} libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel with broadcast communication
highlighting the main differences to Esterel, in particular the stack-based
execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and \emph{C} calls (which behave like in conventional 
imperative languages).

@@

The $mem(id)$ primitive represents all accesses, assignments, and \emph{C} 
function calls that affect a memory location identified by $id$.
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%In accordance with the zero-delay hypothesis of \CEU, $mem$ expressions are 
%considered to be atomic and instantaneous
%
The special notation $nop$ is used to represent an innocuous $mem$ expression 
(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
unused identifier).
%
Note that $mem$ and $await$/$emit$ expressions do not share identifiers, i.e., 
an identifier is either a variable or an event.

@@

This phase detects inconsistencies in \CEU programs, such as unbounded loops, 
suspicious accesses to shared memory, and also ``classical'' semantic analysis, 
such as building a symbol table for checking variable declarations.
%
This phase outputs code in \emph{C}, given the way \CEU is tied to \emph{C} by 
design.
%
Some type checking is delayed to the last phase to take advantage of 
\code{gcc}'s error handling.
Therefore, we annotate the \emph{C} output file with \code{\#line} pragmas 
matching the original file in \CEU.

@@

The final phase packs the generated \emph{C} file with the \CEU runtime and 
platform-dependent functionality, compiling them with \emph{gcc} and generating 
the final binary.
%
The \CEU runtime comprehends the scheduler, timer management, and the external 
\emph{C} API.
%
The platform files include libraries for I/O and bindings to invoke the \CEU 
scheduler on external events.

@@

A node inherits the set $I$ from its direct parent and calculates $O$ according 
to its type:
%
\begin{itemize}
%
\item Nodes that represent expressions, assignments, \emph{C} calls, and 
declarations simply reproduce $O=I$, as they do not await;
%
\item An \code{await e} statement has $O=\{e\}$.
%
\item A \code{break} statement has $O=\{\}$ as it escapes the innermost 
\code{loop} and never terminates, i.e., never proceeds to the statement 
immediately following it (see also \code{loop} below);
%
\item A \emph{sequence node (;)} modifies each of its children to have 
$I_n=O_{n-1}$.
The first child inherits $I$ from its parent node, and the set $O$ for the 
sequence node is copied from its last child, i.e., $O=O_n$.
%
\item A \code{loop} node includes its body's $O$ on its own $I$ ($I=I \cup 
O_{body}$), as the loop is also reached from its own body.
The union of all \code{break} statements' $O$ forms the set $O$ for a 
\code{loop}.
%
\item An \code{if} node has $O=O_{true} \cup O_{false}$.
%
\item A parallel composition may terminate from any of its branches, hence $O = 
O_1 \cup ... \cup O_n$.
\end{itemize}

@@

\CEU favors a fine-grained use of trails, being common the use of trails that 
await a single event.
For this reason, \CEU{} does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and runtime flags.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%
Translating this idea to \emph{C} is straightforward~\cite{wsn.osm,wsn.ocram}: 
memory for blocks in sequence are packed in a \code{struct}, while blocks in 
parallel, in a \code{union}.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at a given time.
A given position in the memory may hold different data (with variable sizes) 
during runtime.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with corresponding 
memory layout.
%
Each variable is assigned a unique $id$ (e.g. \code{a\_1}) so that variables 
with the same name can be distinguished.
%
The \code{do-end} blocks in sequence are packed in a \code{union}, given that 
their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
\code{MEM.b\_2} can safely share the same memory slot.
%
The example also illustrates the presence of runtime flags related to the 
parallel composition, which also reside in reusable slots in the static memory.

@@

In the context of embedded systems, the size of \code{trail\_t} is typically 
only 3 bytes (1 byte for each field), imposing a negligible memory overhead 
even for trails that only await a single event and terminate.
%
For instance, the \emph{CTP} collection protocol ported to \CEU reaches eight 
simultaneous lines of execution with an overhead of 2\% in comparison to the 
original version in \emph{nesC}~\cite{wsn.nesc} (a dialect of \emph{C} for 
event-driven programming)~\cite{ceu.sensys13}.

@@

In the final generated code in \emph{C}, each trail label representing an entry 
point becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in the left of the figure, the compiler extracts the entry 
points and associated trails, e.g., the label \code{Awake\_e} will execute on 
\code{TRAIL-0} (line 7).
%
For each statement that pauses (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail into segments with associated 
entry points.
%
The entry points translate to an \code{enum} in the generated code (lines 
1--10, in the middle of the figure).
The state of trails translate to a vector of type \code{trail\_t} with the 
maximum number of simultaneous trails (lines 12--15).
On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
(line 13), while all others are set to \code{INACTIVE} (in the example, only 
one, in line 14).

@@

\subsection{The External \emph{C} API}

As a reactive language, the execution of a program in \CEU is guided entirely 
by the occurrence of external events.
From the implementation perspective, there are three external sources of input 
into programs, which are all exposed as functions in a \emph{C} API:

@@

\begin{comment}
Note that \CEU{} code running from a call to \code{ceu\_go\_async} may emit an 
input event or the passage of time.
In this case, the \emph{C} implementation makes a tail call to the 
corresponding handler (i.e.  \code{ceu\_go\_event} or \code{ceu\_go\_time}), as 
synchronous code has higher priority.

@@

@@

Esterel has different compilation backends that synthesizes to software and 
also to hardware circuits~\cite{esterel.emperor,esterel.tutorial}.
Among the software-based approaches, \emph{SAXO--RT}~\cite{esterel.saxort} is 
the closest to our implementation with respect to trail allocation and 
scheduling:
The compiler slices programs into ``control points'' (analogous to our ``entry 
points'') and rearranges them into a directed acyclic graph respecting the 
constructive semantics of Esterel.
Then, it flattens the graph into sequential code in \emph{C} suitable for 
static scheduling.

@@

% Bibliography
\bibliographystyle{ACM-Reference-Format-Journals}
\bibliography{my,other}
                             % Sample .bib file with references that match those in
                             % the 'Specifications Document (V1.5)' as well containing
                             % 'legacy' bibs and bibs with 'alternate codings'.
                             % Gerry Murray - March 2012

>>>>

\usepackage{verbatim}
\usepackage{xspace}
\newcommand{\C}{\emph{C}\xspace}
\newcommand{\VM}{\emph{VM}\xspace}
\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\code}[1] {{\small{\texttt{#1}}}}
\newcommand{\MM}[1] {\textcircled{\tiny{\textsf{#1}}}}

@@

An established alternative to \C in the field of embedded systems is the family 
of reactive synchronous languages \cite{rp.twelve}.
%
Two major styles of synchronous languages have evolved:
in the \emph{control}--\emph{imperative} style, programs are structured with 
control flow primitives, such as parallelism, repetition, and preemption;
in the \emph{dataflow}--\emph{declarative} style, programs can be seen as 
graphs of values, in which a change to a value is propagated through its 
dependencies without explicit programming.
%
Among the control-based languages, Esterel~\cite{esterel.ieee91} influenced a 
number of other embedded languages, such as \emph{Reactive-C}~\cite{rp.rc}, 
\emph{OSM}~\cite{wsn.osm}, \emph{Sync-C}~\cite{rp.synchc}, and 
\emph{PRET-C}~\cite{rp.pretc}.

@@

In this work, we present a formal semantics of a small synchronous kernel that 
represents a subset of \CEU covering these new functionalities.
%
We also present an implementation of \CEU with two backends:
one aiming for efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of constrained wireless sensor networks, developing a number of applications, 
protocols, and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C, and observed a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for ROM and RAM)~\cite{ceu.sensys13}.
%
Considering the \VM backend, a simple application that blinks three LEDs 
periodically occupies less than 100 bytes and can be completely transmitted in 
4 radio messages.
In contrast, without the preinstalled \VM with basic \emph{I/O} functionality, 
a \C-compiled version occupies more than 2 kbytes~\cite{ceu.terra}.

The rest of the paper is organized as follows:
Section~\ref{sec.ceu} gives an overview of \CEU, exposing its fundamental 
differences to Esterel.
Section~\ref{sec.sem} presents a formal semantics for the control primitives of 
\CEU.
Section~\ref{sec.impl} presents the \C and \VM implementation backends.
Section~\ref{sec.related} discusses other synchronous languages targeting 
embedded systems.
Section~\ref{sec.conclusion} concludes the paper.

@@

To illustrate threats related to abortion, consider the unsafe example, which 
does not compile, in the left of Figure~\ref{lst.fin}, describing the state 
machine of a data collection protocol for sensor 
networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (line 1) 
represent the external interface of the protocol.
%
The protocol has to transmit a packet every minute, unless it receives a 
\code{RETRANSMIT} request to immediately re-transmit it, or a \code{STOP} 
request to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (line 3);
the other periodically transmits the packet (lines 5--18).
%
The periodic transmission is a loop that starts two other trails (lines 6--17):
one handles the immediate retransmission request (line 7);
the other transmits the packet%
\footnote{
The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
with external \C libraries explicit and are required in \CEU.
}
and waits for a confirmation (lines 9--16).
%
%
The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping, in accordance with the specification.
Note that the call to \code{\_send\_enqueue} is \emph{asynchronous}, handing to 
the radio driver a pointer to the lexically-scoped packet.
The driver makes the transmission in the background, holding the packet until 
it signals the application with the \code{SENDACK} to acknowledge completion.
%
At any time, the client may request a retransmission (line 7), which terminates 
the \code{par/or} (line 6), aborts the ongoing transmission (line 14, if not 
idle), and restarts the loop (line 5).
%
The client may also request to stop the whole protocol at any time (line 3).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver, which will possibly transmit 
corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

The unsafe example in the left of Figure~\ref{lst.fin} does not compile because 
\CEU tracks the interaction of \code{par/or} compositions with local variables 
and stateful \C functions (e.g., device drivers) in order to preserve safe 
abortion of trails~\cite{ceu.sensys13,ceu.mod15}.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
In fact, \CEU enforces the programmer to write a \emph{finalization} clause to 
accompany the stateful \C call.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}) executes automatically on external 
abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel with broadcast communication
highlighting the main differences to Esterel, in particular the stack-based
execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and \C calls (which behave like in conventional 
imperative languages).

@@

The $mem(id)$ primitive represents all accesses, assignments, and \C function 
calls that affect a memory location identified by $id$.
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%In accordance with the zero-delay hypothesis of \CEU, $mem$ expressions are 
%considered to be atomic and instantaneous
%
The special notation $nop$ is used to represent an innocuous $mem$ expression 
(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
unused identifier).
%
Note that $mem$ and $await$/$emit$ expressions do not share identifiers, i.e., 
an identifier is either a variable or an event.

@@

This phase detects inconsistencies in \CEU programs, such as unbounded loops, 
suspicious accesses to shared memory, and also ``classical'' semantic analysis, 
such as building a symbol table for checking variable declarations.
%
This phase outputs code in \C, given the way \CEU is tied to \C by design.
%
Some type checking is delayed to the last phase to take advantage of 
\code{gcc}'s error handling.
Therefore, we annotate the \C output file with \code{\#line} pragmas matching 
the original file in \CEU.

@@

The final phase packs the generated \C file with the \CEU runtime and 
platform-dependent functionality, compiling them with \emph{gcc} and generating 
the final binary.
%
The \CEU runtime comprehends the scheduler, timer management, and the external 
\C API.
%
The platform files include libraries for I/O and bindings to invoke the \CEU 
scheduler on external events.

@@

A node inherits the set $I$ from its direct parent and calculates $O$ according 
to its type:
%
\begin{itemize}
%
\item Nodes that represent expressions, assignments, \C calls, and declarations 
simply reproduce $O=I$, as they do not await;
%
\item An \code{await e} statement has $O=\{e\}$.
%
\item A \code{break} statement has $O=\{\}$ as it escapes the innermost 
\code{loop} and never terminates, i.e., never proceeds to the statement 
immediately following it (see also \code{loop} below);
%
\item A \emph{sequence node (;)} modifies each of its children to have 
$I_n=O_{n-1}$.
The first child inherits $I$ from its parent node, and the set $O$ for the 
sequence node is copied from its last child, i.e., $O=O_n$.
%
\item A \code{loop} node includes its body's $O$ on its own $I$ ($I=I \cup 
O_{body}$), as the loop is also reached from its own body.
The union of all \code{break} statements' $O$ forms the set $O$ for a 
\code{loop}.
%
\item An \code{if} node has $O=O_{true} \cup O_{false}$.
%
\item A parallel composition may terminate from any of its branches, hence $O = 
O_1 \cup ... \cup O_n$.
\end{itemize}

@@

\CEU favors a fine-grained use of trails, being common the use of trails that 
await a single event.
For this reason, \CEU{} does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and runtime flags.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%
Translating this idea to \C is straightforward~\cite{wsn.osm,wsn.ocram}: memory 
for blocks in sequence are packed in a \code{struct}, while blocks in parallel, 
in a \code{union}.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at a given time.
A given position in the memory may hold different data (with variable sizes) 
during runtime.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with corresponding 
memory layout.
%
Each variable is assigned a unique $id$ (e.g. \code{a\_1}) so that variables 
with the same name can be distinguished.
%
The \code{do-end} blocks in sequence are packed in a \code{union}, given that 
their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
\code{MEM.b\_2} can safely share the same memory slot.
%
The example also illustrates the presence of runtime flags related to the 
parallel composition, which also reside in reusable slots in the static memory.

@@

In the context of embedded systems, the size of \code{trail\_t} is typically 
only 3 bytes (1 byte for each field), imposing a negligible memory overhead 
even for trails that only await a single event and terminate.
%
For instance, the \emph{CTP} collection protocol ported to \CEU reaches eight 
simultaneous lines of execution with an overhead of 2\% in comparison to the 
original version in \emph{nesC}~\cite{wsn.nesc} (a dialect of \C for 
event-driven programming)~\cite{ceu.sensys13}.

@@

In the final generated code in \C, each trail label representing an entry point 
becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in the left of the figure, the compiler extracts the entry 
points and associated trails, e.g., the label \code{Awake\_e} will execute on 
\code{TRAIL-0} (line 7).
%
For each statement that pauses (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail into segments with associated 
entry points.
%
The entry points translate to an \code{enum} in the generated code (lines 
1--10, in the middle of the figure).
The state of trails translate to a vector of type \code{trail\_t} with the 
maximum number of simultaneous trails (lines 12--15).
On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
(line 13), while all others are set to \code{INACTIVE} (in the example, only 
one, in line 14).

@@

\subsection{The External \C API}

As a reactive language, the execution of a program in \CEU is guided entirely 
by the occurrence of external events.
From the implementation perspective, there are three external sources of input 
into programs, which are all exposed as functions in a \C API:

@@

\begin{comment}
Note that \CEU{} code running from a call to \code{ceu\_go\_async} may emit an 
input event or the passage of time.
In this case, the \C implementation makes a tail call to the corresponding 
handler (i.e.  \code{ceu\_go\_event} or \code{ceu\_go\_time}), as synchronous 
code has higher priority.

@@

Terra is a system for programming wireless sensor network applications that 
uses \CEU as its scripting language~\cite{ceu.terra}.
%
Figure \ref{fig.terra} shows the three basic elements of Terra:
\CEU as the scripting language,
the set of customized pre-built components,
and the embedded virtual-machine engine which can disseminate and install 
bytecode images dynamically.
%
This approach aims to combine the flexibility of remotely uploading code with 
the expressiveness and safety guarantees of \CEU.

\begin{figure}[!htb]
 \centering
 \includegraphics[width=0.7\columnwidth]{terra}
\caption{Terra programming system basic elements.}
 \label{fig.terra}
\end{figure}

\CEU scripts in Terra cannot execute arbitrary \C code, instead, they rely on 
pre-built components that can be customized for different application domains.
%
Considering the domain of WSNs, Terra already provides components organized in 
four areas: radio communication, group management, data aggregation, and local 
operations (e.g., access to sensors and actuators).
%
When creating an instance of the \VM, the programmer can choose whether or not 
to include each component, setting different abstraction boundaries.
%
The generated \VM has to be preloaded into the embedded devices before they are 
physically distributed.

The communication between \CEU and the Terra components is mostly through 
events:
scripts make requests through \code{output} events and \code{await} answers 
through \code{input} events.
%
Terra also provides system calls for initialization and configuration of 
components.

TODO: exemplo de implementacao/uso de um componente
- Terra: binding em nesC
- Ceu: header de configuracao
- (analogo a Figura 19)

Considering the compilation process illustrated in Figure~\ref{fig.impl} for 
the \C backend, the main difference resides in the \emph{Code Generation}, 
which now outputs assembly instructions for the Terra \VM, instead of \C.

TODO: falar um pouco do assembly/bytecode, quais as decisoes envolvidas
- (analogo a Figura 18)

\begin{comment}
Instead of generating code in C, Terra generates the vmt assembly code using 
the original Céu labels for trails entry points.
In the final phase for code generation, Terra compiler generates the vmt 
bytecode file using the program address in place of Céu labels.
The use of address allows the Céu scheduler to execute directly a trail without 
a index table to replace the switch-case.
Related to memory layout and trail allocation, Terra uses the same data 
organization approach for all Céu controls and variables.
All Céu data and control structures for a specific program are determined at 
compile time.
Terra arranges all these structures in the beginning of virtual machine memory 
space.
The original Céu run time control functions are maintained with a small 
modification where accesses to control structures have an off-set relative to 
the begin of the virtual machine memory space.
\end{comment}

@@

Esterel has different compilation backends that synthesizes to software and 
also to hardware circuits~\cite{esterel.emperor,esterel.tutorial}.
Among the software-based approaches, \emph{SAXO--RT}~\cite{esterel.saxort} is 
the closest to our implementation with respect to trail allocation and 
scheduling:
The compiler slices programs into ``control points'' (analogous to our ``entry 
points'') and rearranges them into a directed acyclic graph respecting the 
constructive semantics of Esterel.
Then, it flattens the graph into sequential code in \C suitable for static 
scheduling.

@@

% Bibliography
\bibliographystyle{ACM-Reference-Format-Journals}
\bibliography{my,other,terra}
                             % Sample .bib file with references that match those in
                             % the 'Specifications Document (V1.5)' as well containing
                             % 'legacy' bibs and bibs with 'alternate codings'.
                             % Gerry Murray - March 2012
